---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions, echo = FALSE}
library(knitr)
opts_chunk$set(cache = TRUE)
```


## Loading and preprocessing the data
```{r load}
if (!file.exists("./activity.zip")) {
  fileURL = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileURL, "./activity_data.zip")
  unzip("activity.zip")
}

activity = read.csv("activity.csv")
activity$date = as.Date(activity$date)
summary(activity)
```

## What is mean total number of steps taken per day?
```{r meanTotalSteps}
totalSteps = aggregate(steps ~ date, data = activity, FUN = sum, na.rm = T)
mean(totalSteps$steps)
median(totalSteps$steps)

library(ggplot2)
ggplot(totalSteps, aes(x = steps)) + 
  geom_histogram(colour = "black", fill = "white", binwidth = 1000) +
  geom_vline(aes(xintercept = mean(totalSteps$steps)), color="red", linetype="dashed", size=1) +
  ggtitle("Total number of steps per day") 

```

## What is the average daily activity pattern?
```{r averageDaily}
library(dplyr, warn.conflicts = FALSE)
meanStepsByInterval = activity %>% tbl_df() %>% group_by(interval) %>% summarise(mean_steps = mean(steps, na.rm = TRUE))

ggplot(meanStepsByInterval, aes(x = interval, y = mean_steps)) + geom_line() + ggtitle("Average number of steps by interval")
meanStepsByInterval$interval[which.max(meanStepsByInterval$mean_steps)]
```

## Imputing missing values
```{r missing}
sum(!complete.cases(activity))

activity_complete = activity
indices = which(is.na(activity$steps))

NAs = subset(activity, subset = is.na(steps), select = c("interval", "steps"))
activity_complete$steps[indices] = meanStepsByInterval$mean_steps[match(activity$interval[indices], meanStepsByInterval$interval)]

totalStepsComplete = aggregate(steps ~ date, data = activity_complete, FUN = sum)
mean(totalStepsComplete$steps)
median(totalStepsComplete$steps)

library(ggplot2)
ggplot(totalStepsComplete, aes(x = steps)) + 
  geom_histogram(colour = "black", fill = "white", binwidth = 1000) +
  geom_vline(aes(xintercept = mean(totalSteps$steps)), color="red", linetype="dashed", size=1) +
  ggtitle("Total number of steps per day") 

```

We can see that there is no effect of imputing on mean and median because we were imputing using the mean values for each of the intervals.

## Are there differences in activity patterns between weekdays and weekends?
```{r weekdays}
activity_complete$typeOfDay = factor(sapply(weekdays(activity_complete$date), 
       function(x) {
         if (x %in% c("Saturday", "Sunday")) return("weekend")
         return("weekday")
       }))

meanStepsByInterval = activity_complete %>% tbl_df() %>% group_by(interval, typeOfDay) %>% summarise(mean_steps = mean(steps, na.rm = TRUE))

ggplot(meanStepsByInterval, aes(x = interval, y = mean_steps)) + 
  facet_grid(facets = typeOfDay ~ .) + 
  geom_line() + 
  ggtitle("Average number of steps by interval")
ggplot(meanStepsByInterval, aes(x = interval, y = mean_steps, color = typeOfDay)) + 
  geom_line() + 
  ggtitle("Average number of steps by interval")

```
